Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-12-04 ms/review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data''.
We thank the AE and revewiers for constructive comments,
which we have addressed in detail, below.
The main changes we have made have to do with our choice of model.
We have added some more discussion
of the biological motivation for the particular model of continuous space we use
(with references to the ecological literature, where it appears) to the Methods (around \llname{ecolcites}),
and in the Discussion (at \llname{demog_disco}).
We have included more explanation of some of the more opaque consequences
of a more realistic demographic model, around \llname{censusscaling}.
Finally, we have added a substantial comparison to the discrete stepping-stone model,
most of which is in the Appendix.

@@

We thank the reviewer for their very constructive comments!
Responses follow below:

\begin{point}{}
    When neighbourhood size is small (10-100), the mean number of IBS tracts ${}>2$bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased. What could be the source of this discrepancy? How exactly mean(IBS) was calculated?
\end{point}

\reply{
    The mean IBS is simply the mean of the vector of lengths of IBS tracts \revref.
    The short answer is that if one splits a chromosome up into more pieces (increasing nIBS),
    the mean length of those pieces must necessarily be smaller.
    At low neighborhood sizes, spatially distant individuals tend to have longer coalescence times,
    leading to more (and hence smaller) IBS tracts.
    This can be seen in Supplemental Figure \ref{fig:allsumstats}.
    Note that skew(IBS) and number of long ($>$1e6) IBS tracts reflect other aspects of the distribution.
}

@@

\begin{point}{}
    Related to the first point, to what extent are the results qualitatively different to those that would be obtained in a stepping-stone model? My interpretation is that they are actually very similar, but I didn't see whether that was explicitly discussed. In some sense, it's still easier to do large simulations in a stepping-stone model so it would be nice to be reassured that that's still ok.
\end{point}

\reply{
    We have added an appendix comparing our model to a reverse-time stepping stone simulation on several relevant summary statistics \llname{apx:stepping_stone}.
    For some of these statistics (e.g., $\theta_\pi$),
    the behavior of the stepping stone model becomes more similar (although not identical)
    to the continuous model as the resolution of the landscape increases,
    but other statistics are more strongly affected by discretization.
    We are also curious how all the other analyses we perform would be affected using other simulation schemes, but hope to explore that aspect in future work. 
    Furthermore, discrete-deme coalescent simulations are certainly faster,
    but come with significant issues either when neighborhood size is lower than the population of a deme,
    or when demes are small enough that the sample size approaches local $N_e$.
}

\begin{point}{}
    The source of equation \eqref{eqn:p_survival} is not obvious to me. I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.
\end{point}

\reply{
    Additional discussion of the model is now provided in the appendix (see ``Demographic model'') \llname{apx:demographic_model}
}

@@

\reply{
    We agree that this is hard to intuit: there are a lot of factors at work.
    We have added some more discussion of these phenomena \llname{censusscaling}.
}

\begin{point}{}
    \autoref{fig:gwas}D: I am surprised by the extent to which the observed values of $-\log_{10}(p)$ fall below the $y=x$ line. Particularly in the lower right panel for large neighbourhood sizes. I would expect that to be close to panmictic - why are the P-values underdispersed? That seems like a potential bug, or else something weird is going on.
\end{point}

\reply{
    We have checked the code to the best of our abilities and did not find a bug causing the underdispersion, but we have switched our visualization in Figure 8D to now show a subset of qqplots as lines, rather than the hexagonal binning over all simulations. This lets us see the density in addition to the range of values, and shows that relatively few points fall below the 1:1 lines. Those that do seem to reflect overcorrection in the regression when using PC coordinates as covariates -- the PCA is capturing some information about the spatial genetic variation which itself covaries only weakly with the phenotype, and as a result we see anomalously low $-log10(p)$ when regressing genotype against phenotype. 
}

\begin{point}{\revref}
    It might be worth citing \citep{haworth2019apparent},
    who do the proposed test (GWAS for birth location) in UK Biobank to illustrate the population structure.
\end{point}

@@

    We definately think that a comparison to the spatial Lambda-Fleming-Viot
    would be an extremely useful thing to do, for the reasons the reviewer mentions,
    but this is beyond the scope of the current study.
    The SLFV model is only known to be an approximation of one particular biologically explicit model
    (of patchy extinction-recolonization)
    and is only conjectured to be a good approximation
    for spatial models more generally.
    To reiterate, the definition of the SLFV was explicitly motivated by a desire to approximate
    spatial models, especially Bolker-Pacala models like ours.
    Therefore, our study does important preliminary work
    for simulation study of the spatial Lambda-Fleming-Viot,
    since it works through some of the issues in simulating biologically realistic spatial models,
    and describes some of the patterns.

\begin{point}{\autoref{fig:genparams}:}
    I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model. For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring. From [the paragraph at \revref], it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges. I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects. If that is the case, then the authors should state it clearly and try to justify it from a biological perspective.
\end{point}

\reply{
    It's a good point that neighborhood size, as defined to be ``number of possible parents'',
    should indeed be the census size.
    However, we are using $N_W$ defined in the same way for both spatial and random mating simulations
    as a way of comparing spatial to ``nonspatial'' models
    keeping other aspects of demography (almost) the same.
    We've added a note clarifying this, at \revref{} and \llname{randomNW}.
    This is indeed part of what's going on.
    We have added some discussion of the census size scaling \llname{censusscaling},
    but haven't added a detailed discussion of whether or not most species experience
    decreased habitat quality at range edges,
    since that seems like it is getting too far into the weeds.
}

@@

\reply{
    Thank you for this suggestion. We have slightly revised the introduction and hope it is now clearer; however since we want to cover a little history and motivation for our continuous model vs stepping-stone approaches in the intro it does have a different flow from the discussion.

}

\begin{point}{}
    I agree that most modern work describes structure as discrete populations connected by migration. However, some methods/studies have explicitly modeled spatial structure, e.g. especially in ecology or using methods like dadi (diffusion approximations). Highlighting some examples of previously identified structure not possible to infer without modeling geography would be helpful to contextualize this work.
\end{point}

\reply{
   We have expanded our citations of some of the relevant ecology literature \llname{ecolcites}, which we hope helps to contextualize the study better.
   We haven't cited dadi, which works with discrete populations
   (the ``diffusion'' is in frequency space).
}

@@

\begin{point}{}
    There is some reference to spatial models using grids (e.g. Rousset 1997). Some additional discussion contextualizing more recent methods like EEMS that also construct demes and model migration through divergence between neighboring demes would be helpful and interesting.
\end{point}

\reply{
    We agree that not noting EEMS was an oversight.
    We have added the most recent EEMS paper to the citations at \revref,
    and have added an Appendix giving a more thorough comparison with stepping-stone models. 
}

@@

\reply{
    We now discuss haplotype methods in the relevant discussion section \llname{ibdnesxn}. However though these methods should be more accurate for recent events it is not clear that this will improve performance per se. The dips in recent inferred Ne from stairwayplot are not just prediction noise, but actually reflect an underlying genealogy in which terminal branches are shorter than expected from a constant-size random-mating population (see e.g. figure \ref{fig:sumstats}A and \ref{fig:ibs}). The interpretation error is that these short branches are generated by spatial structure rather than changes in population size over time -- a point also made in the \citep{Mazet2015} paper we discuss in the introduction and discussion. 
}

\begin{point}{}
    GWAS mixed models. To what extent can spatial signals (e.g. corner, patchy) be corrected with mixed models, e.g. with PCs and PC-adjusted GRM as in Conomos et al, 2016 using PC-AiR and PC-Relate)? Is patchiness related to dispersal? I'm curious how this relates to the predictive ability of GWAS phenotypes with some spatial association that may or may not be associated with environmental effects.
\end{point}

\reply{
    Good question -- we are also interested to know how mixed models perform here, but think that to properly test that we would want to change our design to generate phenotypes from simulated genotypes. This would allow us to evaluate false-negatives in addition to false-positives. This is important because, if mixed models do provide stronger control for stratification they are also likely to remove true signals of causal SNPs if those SNPs covary with spatial structure. We now point to these methods explicitly in the discussion \llname{mixedmodelcallout}, but think that incorporating that study here would make this paper too long and we plan to explore this timely issue in the future. We also think the PC results are still quite relevant as the method is quite population in current studies. 
}

\begin{point}{}
    Code availability. This github link doesn't work, but is important to be able to evaluate for review: \url{https://github.com/petrelharp/spaceness}.
\end{point}

\reply{
Apologies, it was accidentally set to private. The link should work now
(and has been replaced by the primary link, \url{https://github.com/cjbattey/spaceness}).
}

\begin{point}{}
    Definitions and interpretations. There are quite a large number of metrics discussed in Figure \ref{fig:sumstats}, and it's a lot to take in. It might be helpful to have a table with a reminder of what the metric is, its interpretation, and how it is computed.
\end{point}

\reply{
    We have included a table describing the summary statistics in \autoref{table:sumstats}.
}

\begin{point}{\revref}
    Notation: ``Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates. Each offspring is produced with a mate selected randomly from those within distance $3\sigma$, with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2/2\sigma^2)$.'' I think x may be overloaded here, or I'm confused. Clarify?
\end{point}

\reply{
	We have clarified the notation for the second instance (referring to the distance among individuals) to $d$. \revref
}

@@

\begin{point}{}
    When introducing the ``spatial model'' as opposed to this ``random model,'' the more concrete illustration in Figure 1 hasn't yet been referenced, which makes it harder to follow. It would be helpful to introduce this figure with the model. Additionally, when Figure 1 is introduced, the order is from right to left (random, then point, then midpoint). It would be helpful to rearrange the figure to mirror what's in the text.
\end{point}

\reply{
We have rearranged the figure as suggested.

}

\begin{point}{\revref}
    Not sure I follow this example: ``Concretely, an individual at position (x, y) in a 50 × 50 landscape has mean phenotype 100 + 2x/5.''
\end{point}

\reply{
	We have clarified by switching to ``the phenotype $p$ for an individual at location (x,y) is then $p=100+2x/5$''. \revref
}

### 2019-12-05 ms/spaceness.tex
fine-toothed comb pass

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004septimescales}), \revpoint{AE}{1}
much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2018populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}. \revpoint{4}{3}
For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

@@

% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation by Wright \citep{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

@@

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to achieve this \citep{bolker2003spatial}.
In our simulations,
we decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions,
and produces total census sizes that fluctuate around an equilibrium. \revpoint{3}{2}
This also prevents the population clumping seen by \citet{Felsenstein1975}
(Supplemental \autoref{fig:wm_v_dd})). \revpoint{3}{1}
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We simulated populations using the non-Wright-Fisher module in the program SLiM v3.1 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance.

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates. \revpoint{4}{8}
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
the probability of survival until the next time step for individual $i$ is
\begin{equation} \label{eqn:p_survival}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
and the mean lifetime is around $L$;
for more description see Appendix \ref{apx:demographic_model}.

@@

We stored the full genealogy and recombination history of final-generation individuals as tree sequences \citep{Kelleher2018},
as implemented in SLiM \citep{haller2019treesequence}. 
Scripts for figures and analyses are available at \url{https://github.com/petrelharp/spaceness}.

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time. \revpoint{2}{5}
We verify that this procedure produces the same site frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs},
as predicted by theory \citep{ralph2019efficiently}.
Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

Our demographic model includes parameters that control population density ($K$), \revpoint{1}{2}
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations,
and used mean population density ($\rho$, census size divided by total area)
to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$. \revpoint{3}{6}
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.

@@

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well as an isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance analogous to \cite{Rousset1997}'s approach, 
which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
defined to be the set of distances between adjacent sites that differ between the two chromosomes
of each diploid. \revpoint{1}{1}
The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/cjbattey/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}. For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (${}>48$ map units) and nearby (${}<2$ map units) individuals. Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance among individuals carrying each derived allele in a set of simulations representing a range of neighborhood sizes.

@@

To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy.  We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model and plotting these against neighborhood size.

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero \revpoint{1}{5}
to a mean that may depend on spatial position such that the mean of spatially-correlated phenotypes varies by two standard deviations across the landscape.\llabel{phenotypescalingsd}
We then adjust the geographic pattern of mean phenotype to create four spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, the phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. \revpoint{4}{10}
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

\section{Data Availability}
Scripts used for all analyses and figures are available at \url{https://github.com/petrelharp/spaceness}.

>>>>

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. However, mates tend to live near to one another and to their offspring, leading to a positive correlation between genetic differentiation and geographic distance. This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}.
Despite a long history of analytical work describing the genetics of populations distributed across continuous geography
(e.g., \citet{Wright1943,Rousset1997,barton2002neutral,Barton2010,Ringbauer2017,robledoarnuncio2010isolation,Wilkins2002linearcoalescent,Wilkins2004separation}), \revpoint{AE}{1}
much modern work still describes geographic structure as a set of discrete populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical,rousset2011likelihoodbased,Shirk2014,lundgren2019populations} or as an average over such discrete models \citep{Petkova2015,al2019estimating}. \revpoint{4}{3}
For this reason, most population genetics statistics are interpreted with reference to discrete, well-mixed populations, and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} with methods that assume these are randomly mating units.

@@

% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

@@

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations
across continuous geographical space.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to achieve this \citep{bolker2003spatial}.
In our simulations,
each individual's probability of survival is a decreasing function of local population density,
which shifts reproductive output towards low-density regions,
and produces total census sizes that fluctuate around an equilibrium. \revpoint{3}{2}
This also prevents the population clumping seen by \citet{Felsenstein1975}
(Supplemental \autoref{fig:wm_v_dd})). \revpoint{3}{1}
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We simulated populations using the program SLiM v3.1 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ is equal to the mean parent-offspring distance.

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates. \revpoint{4}{8}
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
the probability of survival until the next time step for individual $i$ is
\begin{equation} \label{eqn:p_survival}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
and the mean lifetime is around $L$;
for more description see the Appendix. % \ref{apx:demographic_model}.

@@

We stored the full genealogy and recombination history of final-generation individuals as tree sequences \citep{Kelleher2018},
as implemented in SLiM \citep{haller2019treesequence}. 
Scripts for figures and analyses are available at \url{https://github.com/kern-lab/spaceness}.

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size $10^{8}$bp, 
recombination rate $10^{-9}$ per bp per generation, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time. \revpoint{2}{5}
We verify that this procedure produces the same site frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs},
in agreement with theory \citep{ralph2019efficiently}.
Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

Our demographic model includes parameters that control population density ($K$), \revpoint{1}{2}
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations,
and used mean population density ($\rho$, census size divided by total area)
to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$. \revpoint{3}{6}
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the lifetime total number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period,
% subset the resulting table to include only the last timestep recorded for each individual,
and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citep{pyslim}.

@@

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well (motivated by \citet{Rousset1997}'s results)
the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals. 
% as an isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance analogous to \citet{Rousset1997}'s approach, which we summarized as

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep[e.g.,][]{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
defined to be the set of distances between adjacent sites that differ between the two chromosomes. \revpoint{1}{1}
The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/cjbattey/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons. 
We then calculated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}. For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (more than 48 map units) and nearby (less than 2 map units) individuals. Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance between individuals carrying each derived allele.

@@

To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy.  We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model. % and plotting these against neighborhood size.

@@

In all simulations, the phenotype of each individual
is determined by drawing from a Gaussian distribution with standard deviation 10 \revpoint{1}{5}
and a mean that may depend on spatial position.
In spatially varying models, the mean phenotype differs by two standard deviations across the landscape.
We then adjust the geographic pattern of mean phenotype to create four types of spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, the mean phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$. \revpoint{4}{10}
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype to 120
for individuals with both $x$ and $y$ coordinates below 20 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

\section{Data Availability}
Scripts used for all analyses and figures are available at \url{https://github.com/kern-lab/spaceness}.
